---
title: "Cyclistic Member Type Activity Anlysis"
author: "R Y"
date: '2022-06-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cyclistic Member Type Activity Analysis  
  
### Goal
  
Understand how casual riders and annual members use Cyclistic bikes.   
  
* Similarities and differences  
  
Using these insights, design a new marketing strategy to convert casual riders into annual members.   

### Analysis Results (June 2021 - May 2022)

```{r Anaylses, echo=FALSE, message=FALSE}
library("zoo")
library("lubridate")
library(tidyverse)
library(ggplot2)
library("skimr")
library("janitor")
library(pivottabler)
library("knitr")
library("kableExtra")

# knitr to display dataframes. zoo for as.yearmon

# Load the data in with Original Data set to working directory
setwd("C:/Users/reimi/OneDrive/Google Certificate - R/CaseStudy - Cyclistic/Original Data")
trip202205_df <- read_csv("202205-divvy-tripdata.csv")
trip202204_df <- read_csv("202204-divvy-tripdata.csv")
trip202203_df <- read_csv("202203-divvy-tripdata.csv")
trip202202_df <- read_csv("202202-divvy-tripdata.csv")
trip202201_df <- read_csv("202201-divvy-tripdata.csv")
trip202112_df <- read_csv("202112-divvy-tripdata.csv")
trip202111_df <- read_csv("202111-divvy-tripdata.csv")
trip202110_df <- read_csv("202110-divvy-tripdata.csv")
trip202109_df <- read_csv("202109-divvy-tripdata.csv")
trip202108_df <- read_csv("202108-divvy-tripdata.csv")
trip202107_df <- read_csv("202107-divvy-tripdata.csv")
trip202106_df <- read_csv("202106-divvy-tripdata.csv")

# Change the directory back to where the analysis should be save
setwd("C:/Users/reimi/OneDrive/Google Certificate - R/CaseStudy - Cyclistic")

# stitching all of the dataframes into one
cyclistic_alldata <- rbind(trip202205_df, 
                           trip202204_df,
                           trip202203_df,
                           trip202202_df,
                           trip202201_df,
                           trip202112_df,
                           trip202111_df,
                           trip202110_df,
                           trip202109_df,
                           trip202108_df,
                           trip202107_df,
                           trip202106_df)


# just grab ride_id, rideable_type, started_at, ended_at, and member_casual
# cyclistic_yr_cond <- cyclistic_alldata %>% 
#   select(ride_id, rideable_type, started_date, started_time, ended_date, ended_time, member_casual)
cyclistic_yr_cond <- cyclistic_alldata %>% 
  select(ride_id, rideable_type, started_at, ended_at, member_casual)

# clean column names 
cyclistic_yr_cond <- clean_names(cyclistic_yr_cond)

# remove some of the variables
rm(cyclistic_alldata, trip202106_df, trip202107_df, trip202108_df, trip202109_df, trip202110_df, trip202205_df)

# create ride_length and day_of_week columns
cyclistic_yr_cond$ride_length <- cyclistic_yr_cond$ended_at - cyclistic_yr_cond$started_at
cyclistic_yr_cond$day_of_week <- wday(as.Date(cyclistic_yr_cond$started_at))

# Create day of week key
days_of_week <- c(1:7)
day_of_week <- c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")
days_key <- data.frame(days_of_week, day_of_week)

kbl(days_key, caption = "Days of the Week Key", align = "c") %>% 
  kable_minimal()


# calculate mean/max ride_length and mode of day of week

    # borrow mode function from https://www.statology.org/mode-in-r/
find_mode <- function(x) {
  u <- unique(x)
  tab <- tabulate(match(x, u))
  u[tab == max(tab)]
}

    # by member type
ride_summary_member <- 
  cyclistic_yr_cond %>%
  group_by(member_casual) %>%
  summarise(average_ride_length=mean(ride_length),
            max_ride_length=max(ride_length),
            find_mode(day_of_week)
            )

kbl(ride_summary_member, caption = "Mean/Max of ride length & Popular day of week to ride", align = "c") %>% 
  kable_minimal()

    # making a year-month column
start_dates <- as.POSIXct(cyclistic_yr_cond$started_at, format="%Y/%m/%d")
cyclistic_yr_cond$year_month <- as.yearmon(start_dates)
  #floor_date(start_dates, "month")

    # by month
ride_summary_month <- 
  cyclistic_yr_cond %>%
  group_by(year_month, member_casual) %>%
  summarise(average_ride_length=mean(ride_length),
            max_ride_length=max(ride_length),
            day_of_week_mode=find_mode(day_of_week))

kable(ride_summary_month, caption = "Mean/Max of ride length & Popular day of week to ride by month", align = "c") %>% 
  kable_minimal()

# graph average ride_length per day of the week for member_casual
ride_summary_member_2 <- 
  cyclistic_yr_cond %>%
  group_by(member_casual, day_of_week) %>%
  summarise(average_ride_length=mean(ride_length)/60
  )

kable(ride_summary_member_2, caption = "Average Ride Duration in Min", align = "c") %>% 
  kable_minimal()

# graph day_of_week total counts per member type
ride_summary_rides <- 
  cyclistic_yr_cond %>%
  group_by(member_casual, day_of_week) %>%
  summarise(count_rides=n())

kable(ride_summary_rides, caption = "Total Rides per day of the week by membership type", align = "c") %>% 
  kable_minimal()

```

## Plots of Data Analyzed
  
Graph of average ride length per day of the week for the two member types:  
  
```{r Plots, echo=FALSE, message=FALSE}
ggplot(ride_summary_member_2, aes(x = day_of_week, y = average_ride_length)) +
  geom_col() +
  facet_wrap(~member_casual) +
  labs(title="Average Ride Length per Day of the Week by Member Type",
       subtitle="Day 1=Sunday, 2=Monday, 3=Tuesday, 4=Wednesday, 5=Thursday, 6=Friday, 7=Saturday",
       x="Day of The Week",
       y="Avg Ride Length (min)")
```
  
  
Graph of the total rides per member type throughout the week:    
  
  
```{r plot2, echo=FALSE, message=FALSE}
ggplot(ride_summary_rides, aes(x = day_of_week, y = count_rides)) +
  geom_col() +
  facet_wrap(~member_casual) +
  labs(title="Total rides per Day of the Week by Member Type",
       subtitle="Day 1=Sunday, 2=Monday, 3=Tuesday, 4=Wednesday, 5=Thursday, 6=Friday, 7=Saturday",
       x="Day of The Week",
       y="Total Rides")
```
  
  
Graph of popular ride days per month:  
  
```{r popular rides, echo=FALSE, message=FALSE}
ggplot(ride_summary_month, aes(x = year_month, y = day_of_week_mode, color = member_casual)) +
  geom_point() +
  labs(title="Popular Day of the Week in the Last 12 Months",
       subtitle="Day 1=Sunday, 2=Monday, 3=Tuesday, 4=Wednesday, 5=Thursday, 6=Friday, 7=Saturday",
       x="Year/Month",
       y="Day of The Week")
```
  
  
## Observations/Recommendations for Stakeholders
  
  * Casual riders appear to ride for 2x longer than members ride, though some instances point to a bicycle forgotten to be returned.  
  * Members appear to rent bicycles at a consistent rate through out the week compared to casual riders who rent more often on weekends.  
  * Throughout the year, members consistently ride most often on weekdays compared to casual riders that ride on weekends.  
  * More information on whether members are using the bicycles to commute to work or school will be beneficial for future analyses.